---
title: Finding Related Codes to Diabetes Mellitus (ICD9 250) (with code)
author: J Kyle Armstrong, PhD
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 6
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
---

```{r setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r,  echo=FALSE}
my_billing <- 'sincere-nirvana-175602'
```

# Introduction 

Related to a conversation we had, I thought it would be helpful to showcase an example of how to find codes related to ICD9 250 (Diabetes Mellitus). 

In this example I will utilize the [Observational Health Data Sciences and Informatics](https://www.ohdsi.org/) (OHDSI) [Observational Medical Outcomes Partnership (OMOP) Common Data Model (CDM)](https://www.ohdsi.org/data-standardization/) which is available on Google Big Query.

The first step is to connect to the [OMOP CDM on Google Big Query](https://console.cloud.google.com/bigquery?ws=!1m4!1m3!3m2!1sbigquery-public-data!2scms_synthetic_patient_data_omop)

```{r, message = FALSE}
library(dbplyr)
library(dplyr)
library(bigrquery)
```

```{r}
con <- dbConnect(
  bigquery(),
  project = 'bigquery-public-data',
  dataset = 'cms_synthetic_patient_data_omop',
  billing = my_billing
)
```

Then we will connect to the tables that we will query.
```{r, message = FALSE, warning=FALSE}
person <- tbl(con, "person")
concept <- tbl(con, "concept")
concept_relationship <- tbl(con, 'concept_relationship')
```

# Workflow

Our first step is to find the **concept id** that is related to the "250" **concept code** within the ICD9 vocabulary. In the OMOP CDM each medical code (ICD, HCPCS, NDC, ect) is a concept code and is given a unique primary key in the `concept_id` field within the `concept` table. The concept code is the code native within a particular vocabulary, whereas the concept id is assigned by the OMOP CDM. 
```{r}
query_concept_codes <- c("250")
```

This query will give us the unique `concept_id` that is related to 250 within the ICD9 vocabulary:
```{r}
query_concept_id <- concept %>%
  filter(concept_code %in% query_concept_codes & vocabulary_id == "ICD9CM") %>%
  select(concept_id)
```

The OMOP CDM additionally contains a **standard concept id**, [one concept representing the meaning of each clinical event is designated the Standard](https://ohdsi.github.io/TheBookOfOhdsi/StandardizedVocabularies.html#standardConcepts). We use this query to find the standard concept id for our original code:  
```{r}
standard_concept_id <- concept_relationship %>%
  inner_join(query_concept_id, by= join_by(concept_id_1 == concept_id)) %>%
  filter(relationship_id == 'Maps to') %>%
  select(concept_id_2) 
```


Lastly, now that we have the standard concept id, we can perform another query to find all other concept ids for which this it's standard concept id:
```{r}
result <- concept_relationship %>%
  inner_join(standard_concept_id , by = join_by(concept_id_2 == concept_id_2)) %>%
  filter(relationship_id == 'Maps to') %>%
  select(concept_id_1) %>%
  inner_join(concept, by=join_by(concept_id_1 == concept_id)) %>%
  select(-concept_id_1)
```

# Big Query SQL 
We can generate SQL which can be run on Google Big Query
```{r}
dplyr_to_sql <- function(dplyr_sql){

    output_q <- capture.output(show_query(dplyr_sql), type="output")
    output <- output_q[-1]
    cat(output , sep = "\n")
}
```

```{r, eval=FALSE}
ctes <- "with concept_relationship as (select * from `bigquery-public-data.cms_synthetic_patient_data_omop.concept_relationship`), 
concept as (select * from `bigquery-public-data.cms_synthetic_patient_data_omop.concept`)"

cat(ctes, "\n", "\n")

result %>%
  dplyr_to_sql()
```

We have saved this Google Big Query at <https://console.cloud.google.com/bigquery?sq=361290472179:7e182757d65c42aa97db5e501f3003b7>


# Searchable tables for review  

We can download the results
```{r}
result_local <- result %>%
  collect() %>%
  distinct()
```


Here we create a searchable table of the output, possible options for the `vocabulary_id` include `HCPCS`, `ICD10CM`, `ICD9CM`, and `SNOMED`. 
```{r}
result_local %>%
  arrange(vocabulary_id, concept_code) %>%
  DT::datatable(filter = 'top', 
                options = list(pageLength = 40, autoWidth = TRUE))
```



If we wanted potential related IDC10 codes which should map to ICD9 250 we could provide the following list: 
```{r}
result_local %>%
  filter(vocabulary_id == 'ICD10CM') %>%
  knitr::kable()
```

We can export all the results from this query and send them for review to see if they are applicable. 
```{r}
result_local %>%
  arrange(vocabulary_id, concept_code) %>%
  readr::write_csv(here::here('out/Related_ICD9_250_codes.csv'))
```

# Functionalize queries

Lastly, we would note that we queried a single ICD9 code, it would be relatively easy to edit the definition of `query_concept_codes` in this workflow to account for a list of ICD9 codes or to functionalize this process to get mappings from ICD9 to corresponding ICD10 codes.

```{r}
find_related_OMOP_codes <- function(list_of_codes, vocab){
  query_concept_codes <- list_of_codes
  
  query_concept_id <- concept %>%
    filter(concept_code %in% query_concept_codes & vocabulary_id == vocab) %>%
    select(concept_id)
  
  standard_concept_id <- concept_relationship %>%
    inner_join(query_concept_id, by= join_by(concept_id_1 == concept_id)) %>%
    filter(relationship_id == 'Maps to') %>%
    select(concept_id_2) 
  
  result <- concept_relationship %>%
    inner_join(standard_concept_id , by = join_by(concept_id_2 == concept_id_2)) %>%
    filter(relationship_id == 'Maps to') %>%
    select(concept_id_1) %>%
    inner_join(concept, by=join_by(concept_id_1 == concept_id)) %>%
    select(-concept_id_1)
  
  return(result)
}
```


## Hypertensive Disease Example 

For instance we could apply the function to all ICD9 codes for [Hypertensive Disease](http://www.icd9data.com/2015/Volume1/390-459/401-405/default.htm)

```{r}
Essential_Hypertension <- c('401','401.0','401.1','401.9')
Hypertensive_Heart_Disease <- c('402', '402.0', '402.00', '402.01', 
                                '402.1', '402.10', '402.11', 
                                '402.9', '402.90', '402.91') 
Hypertensive_CKD <- c('403', '403.0', '403.00', '403.01', 
                      '403.1', '403.10', '403.11', 
                      '403.9', '403.90', '403.91')
Hypertensive_Heart_CKD <- c('404', '404.0', '404.00', '404.01', '404.02','404.03', 
                            '404.1', '404.10', '404.11', '404.12', '404.13', 
                            '404.9', '404.90', '404.91', '404.92', '404.93')
Secondary_Hypertension <- c('405', '405.0','405.01','405.09', 
                            '405.1', '405.11', '405.19', 
                            '405.9', '405.91', '405.99')

Hypertensive_Disease <- c(Essential_Hypertension, 
                          Hypertensive_Heart_Disease, 
                          Hypertensive_CKD,
                          Hypertensive_Heart_CKD,
                          Secondary_Hypertension)

Hypertensive_Disease_Related <- find_related_OMOP_codes(Hypertensive_Disease, 'ICD9CM')
```

### Seachable Table for Hypertensive Disease Related codes

Similarly, we can produce searchable outputs 
```{r}
Hypertensive_Disease_Related_local <- Hypertensive_Disease_Related %>%
  collect() %>%
  distinct()

Hypertensive_Disease_Related_local %>%
  arrange(vocabulary_id, concept_code) %>%
  DT::datatable(filter = 'top', 
                options = list(pageLength = 25, autoWidth = TRUE))
```

### Hypertensive Disease IDC10 Related codes
Or give explicit listings or CSVs for review 
```{r}
Hypertensive_Disease_Related_local %>%
  arrange(vocabulary_id, concept_code) %>%
  filter(vocabulary_id == 'ICD10CM') %>%
  knitr::kable()
```
